\section{Estimator Cores}\label{sec:cores}

This section describes the three estimator families that \prismloc\ places
behind a single ROS~2 contract. All three are implemented as ROS-free C++
libraries: the two Monte Carlo localization (MCL) backends and the global
relocalizer live in \texttt{prism\_loc\_core}, and the fusion filter lives in
\texttt{prism\_loc\_fusion}. Consistent with the architectural framing of
Section~\ref{sec:architecture}, the contribution here is factoring, not new
estimation theory: we reuse textbook constructions and report only what the
code implements.

\subsection{Shared Monte Carlo Localization Core}

The two 2D and 3D pose-tracking backends share one particle filter. Following
Thrun-style MCL, the pose posterior at time $t$ is a weighted sample set
$\{(x_t^{[i]}, w_t^{[i]})\}_{i=1}^{N}$ over the planar pose
$x_t = (x, y, \theta)$ (\texttt{Pose2D}), updated by the standard
predict--weight--resample recursion. Only the measurement model differs between
backends, so the entire filter is written against an abstract interface rather
than any specific sensor.

\paragraph{Motion model} \texttt{OdometryMotionModel} implements the
sampling-based odometry model. Each control $u_t$ is decomposed into a first
rotation, a translation, and a second rotation
$(\delta_{r1}, \delta_{t}, \delta_{r2})$ from consecutive odometry poses, and
each particle is advanced by the noise-corrupted increment
\begin{align}
\hat{\delta}_{r1} &= \delta_{r1} - \varepsilon(\alpha_1 \delta_{r1}^2 + \alpha_2 \delta_{t}^2), \nonumber\\
\hat{\delta}_{t}  &= \delta_{t}  - \varepsilon(\alpha_3 \delta_{t}^2 + \alpha_4(\delta_{r1}^2 + \delta_{r2}^2)), \nonumber\\
\hat{\delta}_{r2} &= \delta_{r2} - \varepsilon(\alpha_1 \delta_{r2}^2 + \alpha_2 \delta_{t}^2),
\end{align}
where $\varepsilon(\sigma^2)$ draws zero-mean Gaussian noise with the given
variance and $\alpha_{1\text{--}4}$ (\texttt{MotionParams}) are the usual
rotation/translation noise coefficients. A pure-rotation guard sets
$\delta_{r1}=0$ when the translation is below $10^{-3}$~m to avoid an ill-defined
heading.

\paragraph{Particle filter} \texttt{ParticleFilter} weights each particle by the
measurement model's log-likelihood, subtracting the per-step maximum before
exponentiation for numerical stability, then normalizing. Resampling is
triggered only when the effective sample size
$N_\mathrm{eff} = (\sum_i w^{[i]})^2 / \sum_i (w^{[i]})^2$ drops below
\texttt{resample\_threshold} times $N$, and is KLD-adaptive~\cite{fox2003kld}:
samples are drawn proportional to weight until the count reaches the
Kullback--Leibler bound
\begin{equation}
M_x = \frac{k-1}{2\,\varepsilon_\mathrm{kld}}
      \left(1 - \frac{2}{9(k-1)} + \sqrt{\tfrac{2}{9(k-1)}}\,z_{1-\delta}\right)^{3},
\end{equation}
where $k$ is the number of occupied histogram bins seen so far. Bins are formed
by discretizing $(x,y)$ at \texttt{kld\_bin\_xy} and $\theta$ at
\texttt{kld\_bin\_yaw}; the sample count is clamped to
$[\texttt{min\_particles}, \texttt{max\_particles}]$, and $z_{1-\delta}$
(\texttt{kld\_z}) and $\varepsilon_\mathrm{kld}$ (\texttt{kld\_err}) are the
standard-normal quantile and error bound. The pose estimate is the weighted mean
of $(x,y)$ and the circular mean of $\theta$; the reported $6\times6$ covariance
carries the planar terms and marks $z$, roll, and pitch as unobserved.

\paragraph{Measurement models} A backend supplies a \texttt{MeasurementModel},
whose sole method returns the unnormalized log-likelihood of the current
observation given a candidate pose. Two are provided.
\texttt{Laser2DLikelihoodField} is a Thrun-style likelihood-field model for 2D
scans~\cite{thrun2001robust}: an occupancy grid is preprocessed into a
\texttt{LikelihoodField} that stores, per cell, the Euclidean distance $d$ to the
nearest occupied cell (computed once by a separable distance transform), and each
subsampled beam endpoint contributes
$\log\!\big(z_\mathrm{hit}\,e^{-d^2/2\sigma_\mathrm{hit}^2} + z_\mathrm{rand}/z_\mathrm{max}\big)$,
with \texttt{z\_hit}, \texttt{z\_rand}, \texttt{sigma\_hit}, and
\texttt{max\_beams} as parameters. \texttt{Ndt3DModel} reuses the identical
filter but scores a 3D point cloud against a Normal Distributions Transform
map~\cite{biber2003ndt,magnusson2007scan}. The prior map is voxelized into
per-voxel Gaussians $(\mu_v, \Sigma_v)$ (\texttt{NdtMap}, with small eigenvalues
inflated to a fraction of the largest for numerical conditioning), and the
per-particle likelihood accumulates
$\log\!\big(s(p) + s_\mathrm{floor}\big)$ over cloud points $p$, where
$s(p) = \exp\!\big(-\tfrac{1}{2}(p-\mu_v)^\top \Sigma_v^{-1}(p-\mu_v)\big)$ and
empty voxels score zero. The planar particle state $(x,y,\theta)$ is lifted to a
3D transform at a fixed base height, so the 3D backend estimates the same
$(x,y,\theta)$ as the 2D one. This is the NDT-as-measurement-model (NDT-MCL)
idea: the NDT score usually optimized by Newton registration is instead read out
as a particle likelihood, so the two backends differ only in this class.

\subsection{Branch-and-Bound Global Localization}\label{sec:global}

Recovery from a lost or kidnapped estimate is handled by a deterministic
correlative scan matcher, \texttt{BranchAndBoundMatcher}, following Olson's
multi-resolution correlative search~\cite{olson2009correlative} and the
branch-and-bound formulation of Hess et al.~\cite{hess2016realtime}. At
construction the likelihood field is converted into a probability grid
$p = e^{-d^2/2\sigma_\mathrm{hit}^2}$, and a max-pooling pyramid of
\texttt{max\_depth}$+1$ levels is precomputed, where level $h$ stores the maximum
over a $2^h\times2^h$ cell block. A search then proceeds over discretized poses:
translations within \texttt{linear\_window} and rotations spanning
\texttt{angular\_window} at \texttt{angular\_step} resolution. Coarse nodes are
expanded best-first and pruned whenever their pyramid-derived upper bound cannot
exceed the best exact (leaf) score found so far, which keeps the bound
admissible and the returned pose globally optimal over the discretization. The
result is accepted only when its score reaches \texttt{min\_score\_fraction} of
the number of beams used; an accepted pose re-seeds the particle filter, whereas
a rejected one leaves the current estimate untouched. Unlike the stochastic
random-injection recovery of classical MCL, this search is deterministic and
therefore directly reproducible in a unit test.

\subsection{Error-State Kalman Filter Fusion}

The fusion backend, \texttt{Eskf}, is a 15-state error-state Kalman filter that
follows Sol\`a's quaternion formulation~\cite{sola2017quaternion}. The nominal
state (\texttt{NominalState}) carries position $p$, velocity $v$, orientation
quaternion $q$, accelerometer bias $b_a$, and gyroscope bias $b_g$; the error
state $\delta x \in \mathbb{R}^{15}$ collects
$(\delta p, \delta v, \delta\theta, \delta b_a, \delta b_g)$ with a local
angular error $\delta\theta$. On each IMU sample the nominal state is integrated
in closed form,
\begin{equation}
p \mathrel{+}= v\,\Delta t + \tfrac{1}{2} a_w \Delta t^2,\quad
v \mathrel{+}= a_w \Delta t,\quad
q \leftarrow q \otimes \mathrm{Exp}(\omega\,\Delta t),
\end{equation}
where $a_w = R(q)(a_m - b_a) + g$ and $\omega = \omega_m - b_g$ are the
bias-corrected, gravity-compensated measurements, and $\mathrm{Exp}(\cdot)$ is
the $SO(3)$ exponential (\texttt{so3.hpp}). The error covariance is propagated
with the discrete transition $F$ and additive process noise $Q$ built from the
IMU noise parameters,
\begin{equation}
P \leftarrow F\,P\,F^\top + Q .
\end{equation}
Two measurement updates share the standard Kalman gain
$K = P H^\top (H P H^\top + R)^{-1}$, $\delta x = K\,y$,
$P \leftarrow (I - KH)P$: \texttt{updatePose} applies a 6-DoF NDT pose
observation whose residual stacks the position error and the local orientation
error $\log(q^{-1} q_\mathrm{meas})$, while \texttt{updatePosition} applies a
3-DoF GNSS position observation. GNSS fixes are mapped from geodetic
coordinates into a local ENU frame through a WGS-84 LLA$\rightarrow$ECEF$\rightarrow$ENU
conversion about a fixed datum (\texttt{GeodeticConverter}). After each update
the estimated error is injected into the nominal state --- additively for the
vector components and multiplicatively, $q \leftarrow q \otimes
\mathrm{Exp}(\delta\theta)$, for the orientation --- and the error state is reset
to zero. Because the filter takes plain vectors and quaternions rather than ROS
messages, the entire predict--correct--inject--reset cycle is exercised in ROS-free unit
tests.
