\section{Implementation and Validation}\label{sec:validation}

\prismloc{} is validated at the level of \emph{correctness}, not
field accuracy. Because every estimator is a ROS-free library
(Section~\ref{sec:cores}), each can be exercised by an ordinary
C++ unit test with no middleware runtime, no simulator, and no
recorded bags. The suite comprises 44 GoogleTest cases across four
packages: 25 in the shared \texttt{prism\_loc\_core} (the two MCL
backends and global recovery), 13 in \texttt{prism\_loc\_fusion} (the
ESKF), 4 in the \texttt{prism\_loc} ROS node's adapter layer, and 2 in
\texttt{prism\_loc\_fusion\_ros}.

\textbf{What the cores are pinned to.}
The core tests target the components most likely to regress silently.
The odometry motion model is checked both deterministically
(zero-command and pure-translation cases) and statistically (the mean
of noisy samples approximates the commanded motion). Measurement models
are checked for the monotonicity property they must satisfy to be usable
as a likelihood: the ground-truth pose must out-score a displaced one,
for both the 2D likelihood field and the 3D NDT voxel score. The
particle-filter mechanics---low-variance resampling concentrating on a
heavy particle, and the weighted-mean estimate---are tested directly, as
are the occupancy-grid and NDT-map primitives (world/map round-trips,
out-of-bounds handling, likelihood-field distances, and voxel-mean
recovery). Global recovery is exercised by having branch-and-bound
recover the correct global pose from a single scan---the kidnapped-robot
scenario---and reject an empty scan. Two integration tests close the
loop end to end, driving the full predict--weight--resample cycle until
the estimate converges to ground truth for the \texttt{laser2d} and
\texttt{ndt3d} backends respectively.

\textbf{Fusion and math.}
The ESKF is pinned for both convergence and consistency: a static IMU
leaves the state at rest, constant acceleration reproduces the expected
kinematics, and position and pose updates pull the state toward the
measurement; an integration test tracks a moving target from a
deliberately wrong initial guess. Supporting identities are verified
independently---$SO(3)$ exponential/logarithm round-trips and the
skew/cross-product relation, geodetic ENU round-trips against a datum,
and the \texttt{map}$\to$\texttt{odom} TF composition arithmetic used to
publish the correction transform.

\textbf{Determinism.}
Every stochastic path draws from a single seedable generator exposed as
the ROS parameter \texttt{seed} (default \texttt{42}). The RNG tests
assert that equal seeds reproduce identical sequences and that distinct
seeds differ, so the statistical assertions above are reproducible run
to run rather than flaky.

\textbf{Continuous integration.}
Two CI jobs guard the contract from both sides. A \emph{core} job builds
\texttt{prism\_loc\_core} and \texttt{prism\_loc\_fusion} with plain
CMake and runs \texttt{ctest}, requiring only a compiler and Eigen---no
ROS---which keeps the cores honestly middleware-independent. A separate
\emph{ROS} job builds inside a \texttt{ros:humble} container, resolves
dependencies with \texttt{rosdep}, and runs \texttt{colcon build} and
\texttt{colcon test} across all four packages, so the ROS~2 wrapping is
tested against a real distribution.

\textbf{Scope.}
This validation establishes that the estimators are individually correct
and that all three sit behind one working ROS~2 contract; it is
deliberately not an accuracy benchmark. The planned quantitative field
evaluation is discussed with the other roadmap items in
Section~\ref{sec:limitations}.
